Natural transformations

### 2017-07-26 papers/base-3/paper-3.tex
Fix typo
Add missing $$
Fix some typos

\def\pathToRoot{../../}\input{\pathToRoot headers/header_lecturenotes}
% \input{\pathToRoot headers/theoremenvts.tex}

@@

For example, concider a discrete category $\catA$ and an arbitrary category $\catB$.
Now let $\parpair{\catA}{\catB}{F}{G}$ be functors from $\catA$ to $\catB$.
Now because $\catA$ is discrete, we can ignore the morphism-mappings of $F$ and $G$ (since there are only identity morphisms in $\catA$ that are mapped to identity morphisms in $\catB$) and write $F$ and $G$ as families
of objects in $\catB$
$\bigl(F(A)\bigr)_{A\in\catA}$ and $\bigl(G(A)\bigr)_{A\in\catA}$.
When we have a mapping $\alpha$ of $\catB$-morphisms
$\bigl(F(A)\toby{\alpha_A}G(A)\bigr)_{A\in\catA}$
this $\alpha$ could be regarded as a mapping between the functors $F$ and $G$.
Indeed this construction is a special case of the more general definition of natural transformations.

We first define what a natural transformation in general is precisely:
\begin{defn}[Natural transformation]
  \label{def:nat}
  Let $\catA$ and $\catB$ be categories and $\parpair{\catA}{\catB}{F}{G}$ be two functors from $\catA$ to $\catB$.
  Let $\alpha$ be a family of maps
  $\bigl(F(a)\toby{\alpha_A}G(A)\bigr)_{A\in\catA}$.
  Then we define that $\alpha$ is a \demph{natural transformation} from $F$ to $G$ iff
  for every objects and maps $A, A' \in \catA, f \from A \to A'$, the sqare
  \begin{equation}
    \label{eq:nat}
    \begin{array}{c}
      \xymatrix{
        F(A) \ar[r]^{F(f)} \ar[d]_{\alpha_A} & F(A') \ar[d]^{\alpha_{A'}} \\
        G(A) \ar[r]_{G(f)} & G(A')
      }
    \end{array}
  \end{equation}
  commutes.
  The maps $\alpha_A$ are then called the \demph{components} of $\alpha$.
  The commuting square (\ref{eq:nat}) is also called the \demph{naturality axiom} (throught this is not a axiom in logical sense but a defined property for some $\alpha$).
\end{defn}

@@

\begin{exmp}[Discrete Categories]
  For the first example, the mapping
  $\bigl(F(A)\toby{\alpha_A}G(A)\bigr)_{A\in\catA}$
  the naturality axiom (\ref{eq:nat}) holds, because $\catA$ is discrete, so every morphism $f \from A \to A'$ in $\catA$ is an identy morphism (hence $A=A'$). The diagramm
  \[
    \xymatrix{
      F(A) \ar[r]^{F(I)} \ar[d]_{\alpha_A} & F(A) \ar[d]^{\alpha_{A}} \\
      G(A) \ar[r]_{G(I)} & G(A)
  }\]
  commutes trivially for every identity morphism $I \from A \to A$. Hence $\alpha$ is a natural transformation according to definition~\ref{def:nat}.
\end{exmp}

\begin{exmp}[Determinants and commutative rings]
  Let $n\in\nat$ and $R$ be a commutative ring.
  We have a forgetful functor $U \from \CRing \to \Mon$ by forgetting the addition of the commutative ring.
  Furthermore the set of all $n \times n$ matrices forms a monoid under matrix multiplication.
  If we map $R \in \CRing$ to $R^{n \times n}$ and ring homomorphisms $f \from R \to S$ to
  $\lambda A \in R^{n \times n}. \bigl( f(A_{i,j}) \bigr)_{1 \le i, j \le n}$,
  i.~e. applying $f$ element-wise to the matrix $A$, we get another functor $M_n \from \CRing \to \Mon$.
  We will now see that the \emph{determinant} is a natural transformation between this two functors $M_n$ and $U$.

  First we note that for every ring the determinant of a matrix over this ring is defined canonically.
  Furthermore by definition if $A \in R^{n \times n}$ then $\det A \in R$ for every ring $R$.
  We therefore can establish a family of mappings $\det$:
  $\bigl(M_n(R)\toby{\det_R}U(R)\bigr)_{R\in\CRing}$, because $U(R) = R$.
  Now, to show that $\det$ is indeed a natural transformation, we have to show that the following diagramm commutes for every ring homomorphism $f \from R \to S$:
  \[\xymatrix{
    M_n(R) \ar[r]^{M_n(f)} \ar[d]_{\det_R} & M_n(S) \ar[d]^{\det_S} \\
    U(R) \ar[r]_{U(f)} & U(s)
  }\]
  By unfolding the definitions of $U$ and $M_n$ and applying functional extentionality, this equivalent to
  $f(\det_R(f(A))) = \det_S (\bigl( f(A_{i,j}) \bigr)_{1 \le i, j \le n})$
  for every $A \in R^{n \times n}$.
  This already is a standard result of linear algebra.
  Hence $\det$ is really a natural transformation between the functors $M_n$ and $U$. Or, written as a diagram:
  \[\xymatrix{
    \CRing \rtwocell<5>^{M_n}_{U}{\alpha} & \Mon
  }\]
\end{exmp}

\begin{defn}[Functor Category]
  Natural transformations, as other types of maps that we've encountered earlier
  can be composed. That is, for any two categories $\cat{A}$ and $\cat{B}$,
  functors $F, G, H$ between them, natural transformations $\alpha \from F \to
  G$ and $\beta \from G \to H$:
  \vspace{-2ex}
  \[
  \xymatrix@C+1.5em{
  \cat{A}
  \ruppertwocell<8>^F{\alpha}
  \ar[r]|G
  \rlowertwocell<-8>_H{\beta}
  &
  \cat{B}
  },
  \]
  \vspace{-3ex}\\
  there is a composite natural transformation $\beta \of \alpha$
  \[
  \xymatrix@C+1.5em{
  \cat{A} \rtwocell^F_H{\hspace{.8em}\beta\of\alpha} &\cat{B}
  }
  \]
  defined by $(\beta\of\alpha)_A = \beta_A \of \alpha_A$ for all $A \in \cat{A}$.
  There is also an identity natural transformation
  \[
  \xymatrix@C+1em{
  \cat{A} \rtwocell^F_F{\hspace{.5em}1_F} &\cat{B}
  }
  \]
  on any functor $F$, defined by $(1_F)_A = 1_{F(A)}$.

@@

\begin{proof}
  \begin{itemize}
    \item[($\rightarrow$)]
      By definition of natural isomorphism we have a natural transformation
      $\beta\from G \to F$, such that \[\beta \circ \alpha = 1_{F} \text{ and }
      \alpha \circ \beta = 1_{G}. \] We show that $\forall A \in \cat{A}.\,
      \beta_A \circ \alpha_A = 1_{F(A)} \text{ and } \alpha_A \circ \beta_A =
      1_{G(A)}$.  Suppose $A \in \cat{A}$.  We know by definition of the
      identity natural transformation that $(\beta \circ \alpha)_A = 1_{F(A)}$.
      The claim $\beta_A \circ \alpha_A = 1_{F(A)}$ follows by definition of
      composition of natural transformations.  $\alpha_A \circ \beta_A =
      1_{G(A)}$ is shown analogously.
    \item[($\leftarrow$)]
      We know that for all $A \in \cat{A}$ there is a $\beta_A$ such that
      \[\beta_A \circ \alpha_A = 1_{F(A)} \text{ and } \alpha_A \circ \beta_A =
      1_{G(A)}.\] First we prove that $\beta$ is a natural transformation.
      Suppose $f\from A \to A'$.  We know that $\alpha$ is a natural
      transformation:
      \begin{alignat*}{2}
          & & G(f) \circ \alpha_A & = \alpha_{A'} \circ F(f) \\
          &\Rightarrow &\quad G(f) \circ \alpha_A \circ \beta_A & = \alpha_{A'} \circ F(f) \circ \beta_A\\
          &\Rightarrow & G(f) & = \alpha_{A'} \circ F(f) \circ \beta_A\\
          &\Rightarrow & \beta_{A'} \circ G(f) & = \beta_{A'} \circ \alpha_{A'} \circ F(f) \circ \beta_A\\
          &\Rightarrow & \beta_{A'} \circ G(f) & = F(f) \circ \beta_A
      \end{alignat*}
      By similar reasoning to (a) we get that $\beta$ is the inverse of
      $\alpha$.  Therefore $\alpha$ is a natural isomorpism.
  \end{itemize}
\end{proof}

@@

\section{Equivalence of Categories}
We now want to examine when two categories are the same.
\begin{exmp}
  \[
    \begin{tikzcd}
      \catA: & & \catB:\\
      A \arrow[loop left] & & B \arrow[bend left]{d} \arrow[loop left] \\
                          & & B' \arrow[bend left]{u} \arrow[loop left]
    \end{tikzcd}
  \]
\end{exmp}
todo: explain example

@@

\section{Horizontal Composition}

>>>>

\def\pathToRoot{../../}\input{\pathToRoot headers/header_lecturenotes}

@@

For example, consider a discrete category $\catA$ and an arbitrary category $\catB$.
Now let $\parpair{\catA}{\catB}{F}{G}$ be functors from $\catA$ to $\catB$.
Now because $\catA$ is discrete, we can ignore the morphism-mappings of $F$ and $G$ (since there are only identity morphisms in $\catA$ that are mapped to identity morphisms in $\catB$) and write $F$ and $G$ as families
of objects in $\catB$
$\bigl(F(A)\bigr)_{A\in\catA}$ and $\bigl(G(A)\bigr)_{A\in\catA}$.
When we have a mapping $\alpha$ of $\catB$-morphisms
$\bigl(F(A)\toby{\alpha_A}G(A)\bigr)_{A\in\catA}$
this $\alpha$ could be regarded as a mapping between the functors $F$ and $G$.
Indeed this construction is a special case of the more general definition of natural transformations.

First we define precisely what a natural transformation is in general.
\begin{defn}[Natural transformation]
  \label{def:nat}
  Let $\catA$ and $\catB$ be categories and $\parpair{\catA}{\catB}{F}{G}$ be two functors from $\catA$ to $\catB$.
  Let $\alpha$ be a family of maps
  $\bigl(F(a)\toby{\alpha_A}G(A)\bigr)_{A\in\catA}$.
  Then we define that $\alpha$ is a \demph{natural transformation} from $F$ to $G$ if
  for every objects and maps $A, A' \in \catA, f \from A \to A'$, the square
  \begin{equation}
    \label{eq:nat}
    \begin{array}{c}
      \xymatrix{
        F(A) \ar[r]^{F(f)} \ar[d]_{\alpha_A} & F(A') \ar[d]^{\alpha_{A'}} \\
        G(A) \ar[r]_{G(f)} & G(A')
      }
    \end{array}
  \end{equation}
  commutes.
  The maps $\alpha_A$ are then called the \demph{components} of $\alpha$.
  The commuting square (\ref{eq:nat}) is also called the \demph{naturality axiom} (through this is not a axiom in logical sense but a defined property for some $\alpha$).
\end{defn}

@@

\begin{exmp}[Discrete categories]
  In the first example, the naturality axiom (\ref{eq:nat}) holds for the mapping
  $\bigl(F(A)\toby{\alpha_A}G(A)\bigr)_{A\in\catA}$,
  because $\catA$ is discrete, so every morphism $f \from A \to A'$ in $\catA$ is an identity morphism (hence $A=A'$). The diagram
  \[
    \xymatrix{
      F(A) \ar[r]^{F(I)} \ar[d]_{\alpha_A} & F(A) \ar[d]^{\alpha_{A}} \\
      G(A) \ar[r]_{G(I)} & G(A)
  }\]
  commutes trivially for every identity morphism $I \from A \to A$. Hence $\alpha$ is a natural transformation according to definition~\ref{def:nat}.
\end{exmp}

\begin{exmp}[Determinants and commutative rings]
  Let $n\in\nat$ and $R$ be a commutative ring.
  We have a forgetful functor $U \from \CRing \to \Mon$ by forgetting the addition of the commutative ring.
  Furthermore the set of all $n \times n$ matrices forms a monoid under matrix multiplication.
  If we map $R \in \CRing$ to $R^{n \times n}$ and ring homomorphisms $f \from R \to S$ to
  $\lambda A \in R^{n \times n}. \bigl( f(A_{i,j}) \bigr)_{1 \le i, j \le n}$,
  i.~e. applying $f$ element-wise to the matrix $A$, we get another functor $M_n \from \CRing \to \Mon$.
  We will now see that the \emph{determinant} is a natural transformation between $M_n$ and $U$.

  First we note that for every ring the determinant of a matrix over this ring is defined canonically.
  Furthermore by definition $\det A \in R$ holds for $A \in R^{n \times n}$.
  We therefore can establish a family of mappings $\det$:
  $\bigl(M_n(R)\toby{\det_R}U(R)\bigr)_{R\in\CRing}$, because $U(R) = R$.
  Now, to show that $\det$ is indeed a natural transformation, we have to show that the following diagram commutes for every ring homomorphism $f \from R \to S$:
  \[\xymatrix{
    M_n(R) \ar[r]^{M_n(f)} \ar[d]_{\det_R} & M_n(S) \ar[d]^{\det_S} \\
    U(R) \ar[r]_{U(f)} & U(s)
  }\]
  By unfolding the definitions of $U$ and $M_n$ and applying functional extensionality, this is equivalent to
  $f(\det_R(f(A))) = \det_S (\bigl( f(A_{i,j}) \bigr)_{1 \le i, j \le n})$
  for every $A \in R^{n \times n}$.
  This already is a standard result of linear algebra.
  Hence $\det$ is really a natural transformation between the functors $M_n$ and $U$. Or, written as a diagram:
  \[\xymatrix{
    \CRing \rtwocell<5>^{M_n}_{U}{\hspace{.7em}\det} & \Mon
  }\]
\end{exmp}

\section*{Functor Category}
\begin{defn}[Functor Category]
  Natural transformations, as other types of maps that we have encountered earlier
  can be composed. That is, for any two categories $\cat{A}$ and $\cat{B}$,
  functors $F, G, H$ between them, natural transformations $\alpha \from F \to
  G$ and $\beta \from G \to H$:
  \vspace{-2ex}
  \[
  \xymatrix@C+1.5em{
  \cat{A}
  \ruppertwocell<8>^F{\alpha}
  \ar[r]|G
  \rlowertwocell<-8>_H{\beta}
  &
  \cat{B}
  },
  \]
  \vspace{-3ex}\\
  there is a composite natural transformation $\beta \of \alpha$
  \[
  \xymatrix@C+1.5em{
  \cat{A} \rtwocell^F_H{\hspace{.8em}\beta\of\alpha} &\cat{B}
  }
  \]
  defined by $(\beta\of\alpha)_A = \beta_A \of \alpha_A$ for all $A \in \cat{A}$.
  There is also an identity natural transformation
  \[
  \xymatrix@C+1em{
  \cat{A} \rtwocell^F_F{\hspace{.5em}1_F} &\cat{B}
  }
  \]
  on any functor $F$, defined by $(1_F)_A = 1_{F(A)}$.

@@

\begin{proof}
  \begin{itemize}
    \item[($\rightarrow$)]
      By definition of natural isomorphism we have a natural transformation
      $\beta\from G \to F$, such that \[\beta \circ \alpha = 1_{F} \text{ and }
      \alpha \circ \beta = 1_{G}. \] We show that $\forall A \in \cat{A}.\,
      \beta_A \circ \alpha_A = 1_{F(A)} \text{ and } \alpha_A \circ \beta_A =
      1_{G(A)}$.  Suppose $A \in \cat{A}$.  We know by definition of the
      identity natural transformation that $(\beta \circ \alpha)_A = 1_{F(A)}$.
      The claim $\beta_A \circ \alpha_A = 1_{F(A)}$ follows by definition of
      composition of natural transformations.  $\alpha_A \circ \beta_A =
      1_{G(A)}$ is shown analogously.
    \item[($\leftarrow$)]
      We know that for all $A \in \cat{A}$ there is a $\beta_A$ such that
      \[\beta_A \circ \alpha_A = 1_{F(A)} \text{ and } \alpha_A \circ \beta_A =
      1_{G(A)}.\] First we prove that $\beta$ is a natural transformation.
      Suppose $f\from A \to A'$.  We know that $\alpha$ is a natural
      transformation:
      \begin{alignat*}{2}
          & & G(f) \circ \alpha_A & = \alpha_{A'} \circ F(f) \\
          &\Rightarrow &\quad G(f) \circ \alpha_A \circ \beta_A & = \alpha_{A'} \circ F(f) \circ \beta_A\\
          &\Rightarrow & G(f) & = \alpha_{A'} \circ F(f) \circ \beta_A\\
          &\Rightarrow & \beta_{A'} \circ G(f) & = \beta_{A'} \circ \alpha_{A'} \circ F(f) \circ \beta_A\\
          &\Rightarrow & \beta_{A'} \circ G(f) & = F(f) \circ \beta_A
      \end{alignat*}
      By similar reasoning to ($\leftarrow$) we get that $\beta$ is the inverse
      of $\alpha$.  Therefore $\alpha$ is a natural isomorphism.
  \end{itemize}
\end{proof}

@@

\section*{Equivalence of Categories}
We now want to examine when two categories are the same.
\begin{exmp}
  \[
    \begin{tikzcd}
      \catA: & & \catB:\\
      A \arrow[loop left] & & B \arrow[bend left]{d} \arrow[loop left] \\
                          & & B' \arrow[bend left]{u} \arrow[loop left]
    \end{tikzcd}
  \]
\end{exmp}
todo: explain example

@@

\section*{Horizontal Composition}
